---
title: "TOOR Model: Team OLS Optimized Rating"
author: Zachary P. Malosh, MS, MBA
output: html_notebook
editor_options: 
  chunk_output_type: console
---

\tableofcontents

# Libraries
```{r setup}
library(tidyverse)
```

# Get Games
```{r 'Get Games'}
games <- readr::read_csv('data/afl.csv', col_types = cols(
  Date = col_character(),
  AwayTeam = col_character(),
  AwayScore = col_double(),
  HomeTeam = col_character(),
  HomeScore = col_double(),
  IsNeutralSite = col_logical())
)
games <- data.frame(games, stringsAsFactors = FALSE)
games$IsNeutralSite <- FALSE
games$Date <- lubridate::parse_date_time(games$Date, 'ymd IM p', tz = 'Australia/Melbourne')
games$GameId <- seq_len(nrow(games))
gameCount <- nrow(games)
print(gameCount)
```

Get list of AFL games. Games have the time of the games, the names of the teams involved, and the final score of the game. Additionally, a game ID is generated for each game. There are `r gameCount` games.

# Calculate Observed Values
```{r 'Calculate Observed Values'}
games$GameTotal <- games$AwayScore + games$HomeScore
games$HomeMOV <- games$HomeScore - games$AwayScore # MOV = Margin of Victory
```

For each game, the *GameTotal* (sum of both scores) and *HomeMOV* (Margin Of Victory for the home team) are calculated.

# Get Bradley Terry Ratings
```{r 'Get Bradley Terry Ratings'}
source('R/BradleyTerry.R')
bradleyTerryResults <- bradley_terry(games$GameId, games$HomeTeam, games$AwayTeam, games$HomeScore, games$AwayScore)
bradleyTerryRatings <- bradleyTerryResults$teamStrengths
ratingsDisplay <- data.frame(Team = names(bradleyTerryResults$teamStrengths),
								  Rating = bradleyTerryResults$teamStrengths,
								  stringsAsFactors = FALSE)
rownames(ratingsDisplay) <- NULL
knitr::kable(ratingsDisplay, caption = '2018 AFL Team Ratings')
```

The team ratings, per the Bradley Terry method, can be found above.
 
# Create Minimization Functions
```{r 'Create Min SSE Function'}
source('R/SetupGames.R')

sseMinFunction <- function(homeStrength, awayStrength, homeCoef, awayCoef, homeAdvCoef){
	return(homeAdvCoef + (homeStrength * homeCoef) + (awayStrength * awayCoef))
}

f <- function(p, teamStrengths, games){
	
	# p = PARAMETERS
	# p[1] = Home Field Advantage Coefficient
	# p[2] = Home Strength Coefficient
	# p[3] = Away Strength Coefficient
	
	g <- games %>%
		mutate(HomeStrength = teamStrengths[HomeTeamId],
			   AwayStrength = teamStrengths[AwayTeamId],
			   SseMinResult = sseMinFunction(
			   	homeStrength = HomeStrength,
			   	awayStrength = AwayStrength,
			   	homeAdvCoef = p[1],
			   	homeCoef =  p[2],
			   	awayCoef = p[3]),
			   MarginOfVictoryErrSq = (HomeMarginOfVictory - SseMinResult) ** 2)
	
	result <- sum(g$MarginOfVictoryErrSq)
	return(result)
}
```

# Setup Games
```{r 'Setup Games'}
source('R/SetupGames.R')

g <- setup_games(gameIds = games$GameId,
				 homeTeamIds = games$HomeTeam,
				 awayTeamIds = games$AwayTeam,
				 homeScores = games$HomeScore,
				 awayScores = games$AwayScore,
				 isNeutralSite = FALSE)
```

# Find Coefficients
```{r 'Find Coefficients'}
p <- rep(1, times = 3)
coefficientOptimization <- nlm(f, p, teamStrengths = bradleyTerryRatings, games = g)
coefs <- list(HomeFieldAdv = coefficientOptimization$estimate[1],
			  HomeStr = coefficientOptimization$estimate[2],
			  AwayStr = coefficientOptimization$estimate[3])
print(coefs)
```

# Calculate Estimated Spread
```{r 'Make Predictions'}
g <- g %>%
	mutate(EstimatedSpread = sseMinFunction(homeStrength = as.numeric(bradleyTerryRatings[HomeTeamId]),
											awayStrength = as.numeric(bradleyTerryRatings[AwayTeamId]),
											homeAdvCoef = as.numeric(coefs['HomeFieldAdv']),
											homeCoef = as.numeric(coefs['HomeStr']),
											awayCoef = as.numeric(coefs['AwayStr'])))
```

# Create Logistic Regression for Home Win Probability
```{r 'Create Logistic Regression for Home Win Probability'}
m <- lm(formula = HomeMarginOfVictory ~ EstimatedSpread, data = g)
stdDev <- summary(m)$sigma
g <- g %>%
	mutate(HomeWinProb = 1 - pnorm(0, mean = EstimatedSpread, sd = stdDev),
		   PredictedResult = ifelse(HomeWinProb > 0.5, 1, 0),
		   IsResultPredicted = ifelse(GameResult == PredictedResult, 1, 0),
		   PredictionError = EstimatedSpread - GameResult,
		   ProbErrorSq = (GameResult - PredictedResult) ** 2,
		   LogError = (GameResult * log(HomeWinProb)) + ((1-GameResult) * log(1 - HomeWinProb)))
coefIntercept <- m$coefficients['(Intercept)']
coefSpread <- m$coefficients['EstimatedSpread']
```

# Create Benchmarks
```{r 'Create Benchmarks'}
benchmarks <- data.frame(RawAccuracy = mean(g$IsResultPredicted),
						 RSQ = summary(m)$r.squared,
						 RMSE = sqrt(mean(g$PredictionError ** 2)),
						 MAE = mean(abs(g$PredictionError)),
						 BrierScore = mean(g$ProbErrorSq),
						 LogLoss = -1 * mean(g$LogError))
```

# Create Prediction Functions
```{r 'Create Prediction Functions'}
predictByIds <- function(homeTeamId, awayTeamId, homeSpread = 0){
	homeStrength <- bradleyTerryRatings[as.character(homeTeamId)]
	awayStrength <- bradleyTerryRatings[as.character(awayTeamId)]
	return(predict(homeStrength, awayStrength, homeSpread))
}
predict <- function(homeStrength, awayStrength, homeSpread = 0){
	homeGoalsFavored <- -1 * homeSpread
	awayGoalsFavored <- -1 * homeGoalsFavored
	sseMinResult <- sseMinFunction(homeStrength = homeStrength,
								   awayStrength = awayStrength,
								   homeAdvCoef = as.numeric(coefs['HomeFieldAdv']),
								   homeCoef = as.numeric(coefs['HomeStr']),
								   awayCoef = as.numeric(coefs['AwayStr']))
	predictedHomeSpread <- as.numeric(coefIntercept + (coefSpread * sseMinResult))
	predictedAwaySpread <- -1 * predictedHomeSpread
	homeWinPct <- 1 - pnorm(homeGoalsFavored + ifelse(homeGoalsFavored%%1==0, 0.5, 0), mean = predictedHomeSpread, sd = stdDev)
	awayWinPct <- 1 - pnorm(awayGoalsFavored + ifelse(awayGoalsFavored%%1==0, 0.5, 0), mean = predictedAwaySpread, sd = stdDev)
	drawWinPct <- 1 - (homeWinPct + awayWinPct)
	result <- list(HomeSpread = homeSpread,
				   HomeWinPct = homeWinPct,
				   DrawWinPct = drawWinPct,
				   AwayWinPct = awayWinPct)
	return(result)
}

```
